---
title: "ESM 244 Lab 1"
author: "Rachel Rhodes"
date: "1/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(sf)
library(tmap)

```

### Read in SF trees data
```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"))
```

### Basic wrangling reminders

Refresh some skills for data wrangling & summary statistics using functions in the 'dplyr' package [also see help tab --> cheatsheets --> R markdown for formatting help]

Find the top 5 highest observations of trees by legal status, do some wrangling, make a graph. Not all of these are necessary just to practice.
```{r}
#COUNT function recognizes groups (replayces need for group_by), it als finds counts (replaces n function), and puts in a nice table (replaces summarize) - can also have multiple leves of grouping example: count(legal_status, species)

#DROP_NA - drop any NA values for a variable that you specify

#RELOCATE - allws you to move things around, can specify by location, by class, etc.

top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% 
  slice_max(tree_count, n = 5)
```

Make a graph of those top 5 observations by legal status using ggplot
```{r}
#ggplot defaults to putting them in alphabetical order, you can convert to factor in the wrangling and then do factor reorder OR do it within ggplot FCT_REORDER(variable to reorder, variable to order by)

ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) +
  geom_col() +
  labs(x = "Legal Status", y = "Tree Count") +
  coord_flip() +
  theme_minimal()
```
### A few more data wrangling refresher examples

Only want to keep observations (rows) for Blackwood Acacia trees.

```{r}
#STR_DETECT - look within a variable that we specify and see if it shows up there
#SELECT - keep or exclude columns
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% 
  select(legal_status, date, latitude, longitude)

ggplot(data = blackwood_acacia, aes(x = longitude, y = latitude)) +
  geom_point()
```
### Use tidyr::seperate() and unite() functions

Useful for combining or separating columns

```{r}
#SEPARATE(variable to seperate, into = c("column name", "column name), what is the separator)
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c("spp_scientific", "spp_common"), sep = "::")
```

Example: tidyr::unite()
```{r}
#UNITE("new column name", column1 to unite:column2 unite, what do you wan tto be the separator)
sf_trees_unit <- sf_trees %>% 
  unite("id_status", tree_id:legal_status, sep = "_cool!_" )

```

